---
title: "Advancing public health and data science using INLA - Day 2"
author: "Janet van Niekerk (janet.vanNiekerk@kaust.edu.sa)"
output: html_document
date: "May 2024"
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Users/vanniej/Documents/GitHub/INLA-CDC_2023")

library("INLA")
library("DiagrammeR")
library("car")
library("ggpubr")
library("spdep")
library("RColorBrewer")
library("spatstat")
library("sp")
library("latticeExtra")
library("gridExtra")
library("gstat")
library("raster")
library("ggplot2")
library("ggfortify")
library("survival")
library("joineR")
library("BayesSurvival")
library("icenReg")
library("nloptr")
library("boot")
library("rnaturalearth")
library("leaflet")
```

## Spatial models for areal data

Intro to lattice and irregular lattice

## Example 1 - Besag and Besagproper models
```{r eval=FALSE, echo = TRUE}
inla.doc("besag")
```

In this example we use the North Carolina Sudden Infant Death Syndrome dataset.
```{r Example 1.a}
data(nc.sids)
head(nc.sids)
hist(nc.sids$SID74)
summary(nc.sids)
nc.sids <- st_read(system.file("shapes/sids.shp", package="spData"))

```
We see that the number of deaths is a count variable and therefore propose a Poisson regression model, i.e. 
$$y\sim Poi(E\theta)$$
with $$\theta = \exp(\eta)$$ and $E$ is the expected count. In this way, $\theta$ indicates the risk. We need to calculate the expected count for each observation,
```{r Example 1.b}
p_hat <- sum(nc.sids$SID74) / sum(nc.sids$BIR74)
nc.sids$E74 <- p_hat * nc.sids$BIR74
```
We use the NWBIR as a covariate to investigate if it has any influence on the risk of SIDS. We consider the proportion instead of the count,
```{r Example 1.c}
nc.sids$nwp_hat74 <- nc.sids$NWBIR74 / nc.sids$BIR74
```
Adding a random intercept adds an independent effect, although the counties are probably not independent in terms of socio-economic diversity. So we should rather include a structured random effect such that some "intercepts" are correlated with each other, conveying the dependence in space. Since we have space divided into discrete non-overlapping parts, we can use a Besag or BYM model.
```{r Example 1.d}
#Add besag for counties - a spatial model
#Graph - neighbourhood structure
nc.adj <- poly2nb(nc.sids)
B.nc <- nb2mat(nc.adj, style = "B")

plot(nc.sids$geometry, border="grey")
plot(nc.adj, coords = nc.sids$geometry,  add = TRUE, col = "blue")

#Add covariate to spatial dataset    
nc.sids$CNTY_ID2 <- 1:(length(nc.sids$CNTY_ID))

#Fit the spatial model
result3_sids <- inla(SID74 ~ nwp_hat74 + f(CNTY_ID2, model = "besagproper", graph = B.nc), 
                     data = as.data.frame(nc.sids), 
                     family = "poisson",
                     control.predictor = list(compute = TRUE),
                     E = E74)
```
The posterior summary is
```{r Example 1.e}
summary(result3_sids)
```
The two extra hyperparameters from the besagproper effect are the marginal standard deviation, $\sigma$ and the properness parameter, $d$, and their marginal posteriors are
```{r Example 1.f}
res3_sd_bym <- inla.tmarginal(function(x) sqrt(1/x), result3_sids$marginals.hyperpar$`Precision for CNTY_ID2`)
plot(res3_sd_bym, type = "l", xlab = expression(sigma), ylab = expression(paste("f(", sigma ,"|y)")))

plot(result3_sids$marginals.hyperpar$`Diagonal for CNTY_ID2`, type = "l", xlab = "d", ylab = "f(d|y)")

```
We can again visually see the fit of the predictions from the model to the data,
```{r Example 1.ga}
nc.sids$fit_besag <- result3_sids$summary.fitted.values$mean*nc.sids$E74
plot(nc.sids["SID74"], breaks = seq(0,50, by = 5), main = "Observed counts")
plot(nc.sids["fit_besag"], breaks = seq(0,50, by = 5), main = "Predicted counts")

plot(nc.sids$SID74, ylab = "SIDS Counts")
points(nc.sids$E74*result3_sids$summary.fitted.values[,1], col = "purple", pch = 8, cex = 0.8)
```
When we compare the fit between models we can look at the marginal log-likelihoods, DIC or WAIC.
```{r Example 1.h}
result1_sids <- inla(SID74 ~ nwp_hat74, data = nc.sids, 
                     family = "poisson",
                     control.compute = list(dic = TRUE, waic = TRUE),
                     control.predictor = list(compute = TRUE),
                     E = E74)

result2_sids <- inla(SID74 ~ nwp_hat74 + f(CNTY_ID2, model = "iid"), 
                     data = nc.sids, 
                     family = "poisson",
                     control.compute = list(dic = TRUE, waic = TRUE),
                     control.predictor = list(compute = TRUE),
                     E = E74)

result3_sids <- inla(SID74 ~ nwp_hat74 + f(CNTY_ID2, model = "besagproper", graph = B.nc), 
                     data = as.data.frame(nc.sids), 
                     family = "poisson",
                     control.compute = list(dic = TRUE, waic = TRUE),
                     control.predictor = list(compute = TRUE),
                     E = E74)

print(matrix(rbind(cbind(result1_sids$mlik[1], result1_sids$dic[1], result1_sids$waic[1]),
            cbind(result2_sids$mlik[1], result2_sids$dic[1], result2_sids$waic[1]),
            cbind(result3_sids$mlik[1], result3_sids$dic[1], result3_sids$waic[1])), 
            nrow = 3, ncol = 3,
            dimnames = list(c("Model 1", "Model 2 - iid", "Model 3 - Besag"),
                            c("Marginal log-likelihood", "DIC", "WAIC")
                            )))
```
Let's visualise the fitted values of all the models.
```{r Example 1.g}
nc.sids$fit_besag <- result3_sids$summary.fitted.values$mean*nc.sids$E74
nc.sids$fit_fixed <- result1_sids$summary.fitted.values$mean*nc.sids$E74
nc.sids$fit_iid <- result2_sids$summary.fitted.values$mean*nc.sids$E74
plot(nc.sids["SID74"], breaks = seq(0,50, by = 5), main = "Observed counts")
plot(nc.sids["fit_fixed"], breaks = seq(0,50, by = 5), main = "Predicted counts - fixed")
plot(nc.sids["fit_iid"], breaks = seq(0,50, by = 5), main = "Predicted counts - iid")
plot(nc.sids["fit_besag"], breaks = seq(0,50, by = 5), main = "Predicted counts - besag")


plot(nc.sids$SID74, ylab = "SIDS Counts")
points(nc.sids$E74*result1_sids$summary.fitted.values[,1], col = "blue", pch = 10)
points(nc.sids$E74*result2_sids$summary.fitted.values[,1], col = "red", pch = 19, cex = 0.5)
points(nc.sids$E74*result3_sids$summary.fitted.values[,1], col = "purple", pch = 8, cex = 0.8)
```


## Example 2 - BYM and BYM2 models
```{r eval=FALSE, echo = TRUE}
inla.doc("bym")
inla.doc("bym2")
```
Let's revisit the NC SIDS dataset example.  
Now we add a BYM effect for the different counties,
```{r Example 2.a}
result4_sids <- inla(SID74 ~ nwp_hat74 + f(CNTY_ID2, model = "bym2", graph = B.nc), 
                     data = as.data.frame(nc.sids), 
                     family = "poisson",
                     control.compute = list(dic = TRUE, waic = TRUE),
                     control.predictor = list(compute = TRUE),
                     E = E74)
```
The posterior summary is
```{r Example 2.b}
summary(result4_sids)
```
The two hyperparameters from the BYM effect are the marginal standard deviation, $\sigma$ and the weight parameter, $\phi$, and their marginal posteriors are
```{r Example 2.c}
res4_sd_bym <- inla.tmarginal(function(x) sqrt(1/x), result4_sids$marginals.hyperpar$`Precision for CNTY_ID2`)
plot(res4_sd_bym, type = "l", xlab = expression(sigma), ylab = expression(paste("f(", sigma ,"|y)")))
plot(result4_sids$marginals.hyperpar$`Phi for CNTY_ID2`, type = "l", xlab = expression(phi), ylab = expression(paste("f(", phi ,"|y)")))
```
We can again visually see the fit of the predictions from the model to the data,
```{r Example 2.d}
nc.sids$fit_bym <- result4_sids$summary.fitted.values$mean*nc.sids$E74
plot(nc.sids[c("SID74", "fit_fixed", "fit_iid", "fit_besag", "fit_bym")],
       breaks = seq(0, 50, by = 5),
     main = "Observed and predicted counts")

plot(nc.sids$SID74, ylab = "SIDS Counts")
points(nc.sids$E74*result1_sids$summary.fitted.values[,1], col = "blue", pch = 10)
points(nc.sids$E74*result2_sids$summary.fitted.values[,1], col = "red", pch = 19, cex = 0.5)
points(nc.sids$E74*result3_sids$summary.fitted.values[,1], col = "purple", pch = 8, cex = 0.8)
points(nc.sids$E74*result4_sids$summary.fitted.values[,1], col = "darkgreen", pch = 2, cex = 0.8)
```
We can visualise the spatial and specific effects as well.
```{r Example 2.e}
nc.sids$bym <- result4_sids$summary.random$CNTY_ID2[1:max(nc.sids$CNTY_ID2),"mean"]
plot(nc.sids[c("bym")],
       breaks = seq(-0.4,0.4, by = 0.05),
     main = "BYM component")

nc.sids$bym_besag <- result4_sids$summary.random$CNTY_ID2[1:max(nc.sids$CNTY_ID2)+max(nc.sids$CNTY_ID2),"mean"]
plot(nc.sids[c("bym_besag")],
       breaks = seq(-1,1, by = 0.1),
     main = "Besag component")

nc.sids$bym_iid <- (result4_sids$summary.random$CNTY_ID2[1:max(nc.sids$CNTY_ID2),"mean"]*
  sqrt(result4_sids$summary.hyperpar["Precision for CNTY_ID2", "mean"]) -
  result4_sids$summary.random$CNTY_ID2[1:max(nc.sids$CNTY_ID2)+max(nc.sids$CNTY_ID2),"mean"]*
  sqrt(result4_sids$summary.hyperpar["Phi for CNTY_ID2", "mean"]))/
(1-sqrt(result4_sids$summary.hyperpar["Phi for CNTY_ID2", "mean"]))

plot(nc.sids[c("bym_iid")],
       breaks = seq(-3,3, by = 0.1),
     main = "IID component")

```

When we compare the fit between models we can look at the marginal log-likelihoods, DIC or WAIC.
```{r Example 2.f}
print(matrix(rbind(cbind(result1_sids$mlik[1], result1_sids$dic[1], result1_sids$waic[1]),
            cbind(result2_sids$mlik[1], result2_sids$dic[1], result2_sids$waic[1]),
            cbind(result3_sids$mlik[1], result3_sids$dic[1], result3_sids$waic[1]),
            cbind(result4_sids$mlik[1], result4_sids$dic[1], result4_sids$waic[1])), 
            nrow = 4, ncol = 3,
            dimnames = list(c("Model 1", "Model 2 - iid", "Model 3 - Besag", "Model 4 - BYM"),
                            c("Marginal log-likelihood", "DIC", "WAIC")
                            )))
```
We can also do cross-validation as a means of measuring prediction performance (see https://arxiv.org/pdf/2210.04482).

```{r Example 2.cv}
result1_sids <- inla(SID74 ~ nwp_hat74, data = nc.sids, 
                     family = "poisson",
                     control.compute = list(dic = TRUE, waic = TRUE,
                                            cpo = TRUE, return.marginals = FALSE, control.gcpo = list(enable = TRUE,num.level.sets = 3)),
                     control.predictor = list(compute = TRUE),
                     E = E74)

result2_sids <- inla(SID74 ~ nwp_hat74 + f(CNTY_ID2, model = "iid"), 
                     data = nc.sids, 
                     family = "poisson",
                     control.compute = list(dic = TRUE, waic = TRUE,
                                            cpo = TRUE, return.marginals = FALSE, control.gcpo = list(enable = TRUE,num.level.sets = 3)),
                     control.predictor = list(compute = TRUE),
                     E = E74)

result3_sids <- inla(SID74 ~ nwp_hat74 + f(CNTY_ID2, model = "besagproper", graph = B.nc), 
                     data = as.data.frame(nc.sids), 
                     family = "poisson",
                     control.compute = list(dic = TRUE, waic = TRUE,
                                            cpo = TRUE, return.marginals = FALSE, control.gcpo = list(enable = TRUE,num.level.sets = 3)),
                     control.predictor = list(compute = TRUE),
                     E = E74)

result4_sids <- inla(SID74 ~ nwp_hat74 + f(CNTY_ID2, model = "bym2", graph = B.nc), 
                     data = as.data.frame(nc.sids), 
                     family = "poisson",
                     control.compute = list(dic = TRUE, waic = TRUE,
                                            po = TRUE, return.marginals = FALSE, control.gcpo = list(enable = TRUE,num.level.sets = 3)),
                     control.predictor = list(compute = TRUE),
                     E = E74)

result1_sids$gcpo$groups[[1]]
result2_sids$gcpo$groups[[1]]
result3_sids$gcpo$groups[[1]]
result4_sids$gcpo$groups[[1]]

```

The results for the model selection and validation are:
```{r Example 2.mselect}
print(matrix(rbind(cbind(result1_sids$mlik[1], result1_sids$dic[1], result1_sids$waic[1], mean(result1_sids$gcpo$gcpo)),
            cbind(result2_sids$mlik[1], result2_sids$dic[1], result2_sids$waic[1], mean(result2_sids$gcpo$gcpo)),
            cbind(result3_sids$mlik[1], result3_sids$dic[1], result3_sids$waic[1], mean(result3_sids$gcpo$gcpo)),
            cbind(result4_sids$mlik[1], result4_sids$dic[1], result4_sids$waic[1], mean(result4_sids$gcpo$gcpo))), 
            nrow = 4, ncol = 4,
            dimnames = list(c("Model 1", "Model 2 - iid", "Model 3 - Besag", "Model 4 - BYM"),
                            c("Marginal log-likelihood", "DIC", "WAIC", "G-CV score")
                            )))

```

## Example 3 - Joint disease mapping models (using Besag2 model)
We can do multivariate disease mapping when we take into account the effect of the presence or absence of other diseases simultaneously. These are treated as endogenous variables.
\begin{eqnarray*}
y_{i 1}|\lambda_{i1} &\sim& \text { Poisson }\left(E_{i 1} \lambda_{i 1}\right)\\ 
y_{i 2}|\lambda_{i2} &\sim& \text { Poisson }\left(E_{i 2} \lambda_{i 2}\right) \\
\log (\lambda_{i 1}) &=& m_{1} + \sum_{f = 1}^{F_1} \beta_f X_{i f} + \sum_{r=1}^{R_1} \rho^r(u_{i r}) + b_{i 1} + S_{i}\\
\log (\lambda_{i 2}) &=& m_{2} + \sum_{f = 1}^{F_2} \gamma_f Z_{i f} + \sum_{r=1}^{R_2} \xi^r(v_{i r})+ b_{i 2}  + a \, S_{i}, 
\label{eq:jointmeanmodel}
\end{eqnarray*}
```{r pre-work}
# read data 
merg.data <- readRDS("Data/MAP_data.RDS")

# The map 
map <- ne_countries(returnclass = "sf")

# Extract the map and compile the shape
if (T){
names(map)[names(map) == "iso_a3"] <- "ISO3"
names(map)[names(map) == "name"] <- "NAME"
map <- map[order(map$name_long),]
map$name_long[grepl("Gambia", map$name_long)] <- "Gambia"
map$name_long[map$name_long == "Republic of the Congo"] <- "Congo"

map <- map[order(map$name_long),]

# Select only the common countries from the map
sub_map <- match(map$name_long , merg.data$Common.countries)
map$cc <- sub_map
map_2 <- subset(map,map$cc!="NA" )
map_2 <- map_2[order(map_2$name_long),]

map_2$D.E = merg.data$D.E
map_2$M.E = merg.data$M.E
map_2$M.cases = merg.data$M.cases
map_2$D.cases = merg.data$D.cases
map_2$M.pop = merg.data$M.pop
map_2$D.pop = merg.data$D.pop


library(cleangeo)
rr <- clgeo_CollectionReport(map_2)
summary(rr)
issues <- rr[rr$type == NA,]
map_2_c <- clgeo_Clean(map_2)
}

# Neighbourhood graph
nb <- poly2nb(map_2_c)
nb2INLA("Day 2/map.adj",nb = nb)
g <- inla.read.graph(filename = "Day 2/map.adj")
map_2$ID = 1:length(map_2_c$level)

par(mar = c(0,0,0,0))
plot(map$geometry[map$continent == "Africa"], border = "grey")
plot(map_2_c, add = TRUE)
text(coordinates(map_2_c)[,1], coordinates(map_2_c)[,2], label = map_2$ID)
plot(nb, coords = map_2_c,  add = TRUE, col = "blue", pch = 1, cex = 0.01)

#Edit the graph to manually connect 12 and 14 & 15 and 8
if (F){
ind = c(12,14,15,8)
g1 = g
g1$nnbs[ind]=g1$nnbs[ind] + 1
g1$nbs[[12]] = c(g1$nbs[[12]], 14)
g1$nbs[[14]] = c(g1$nbs[[14]], 12)
g1$nbs[[15]] = c(g1$nbs[[15]], 8)
g1$nbs[[8]] = c(g1$nbs[[8]], 15)
g1$cc$id = rep(1, g1$n)
g1$cc$n = 1
g1$cc$nodes = 1:g1$n
g1$cc$mean = rep(1, g1$n)
  
plot(g)
dev.off()
plot(g1)
}
#OR edit the neighbourhood
if (F){
  nb1 <- nb
  nb1[[12]] <- as.integer(c(14))
  nb1[[14]] <- as.integer(c(nb[[14]], 12), length = length(nb[[14]])+1)
  nb1[[8]] <- as.integer(c(nb[[8]], 15), length = length(nb[[8]])+1)
  nb1[[15]] <- as.integer(c(nb[[15]], 8), length = length(nb[[15]])+1)

  par(mar = c(0,0,0,0))
plot(map$geometry[map$continent == "Africa"], border = "grey")
plot(map_2_c, add = TRUE)
text(coordinates(map_2_c)[,1], coordinates(map_2_c)[,2], label = map_2_c$ID)
plot(nb1, coords = map_2_c,  add = TRUE, col = "blue", pch = 1, cex = 0.01)

nb2INLA("Day 2/map1.adj",nb = nb1)
g1_alt <- inla.read.graph(filename = "Day 2/map1.adj")

plot(g1)
dev.off()
plot(g1_alt)
  }


```
Now we can model Malaria and G6PD mean incidence jointly by sharing a common spatial field using *besag2* (see https://doi.org/10.1098/rsos.230851).
```{r}
b = length(merg.data$M.cases)
y1 = c(merg.data$M.cases , rep(NA,b))
y2 = c(rep(NA,b) , merg.data$D.cases)

E1 = c(merg.data$M.E,rep(NA,b))
E2 = c(rep(NA,b) , merg.data$D.E)

yy = cbind(y1,y2)
EE = cbind(E1, E2)
mu = c(rep(1, b) , rep(2,b))
b1 = c(1:b,rep(NA ,b))
b2 = c(rep(NA,b),1:b)

ID = c(1:b , rep(NA ,b))

ID_copy = c(rep(NA , b), 1:b + b)

m = as.factor(mu)
d = data.frame(yy, m, ID, ID_copy, b1, b2)  

formula = yy ~ -1 + m + offset(log(E1))+ offset(log(E2))+
  f(ID, model = "besag2", graph=g, scale.model = T)+
  f(ID_copy, copy="ID", hyper = list(beta = list(fixed = TRUE)))+
  f(b1 , model = "bym2", graph=g, scale.model=TRUE)+
  f(b2, model = "bym2", graph=g, scale.model=TRUE)

r1 <- inla(formula,
           family = c("poisson","poisson"),
           data = d,
           verbose = F,
           control.predictor = list(compute = T))

round(r1$summary.fixed[,c(1:3,5)],3)

round(r1$summary.hyperpar[,c(1,3,5,6)],3)
```
We can plot the relative risk from the model.
```{r}
#Fitted values
map_2$fit_besag2_m <- r1$summary.fitted.values[1:b,"mean"]/map_2$M.E
map_2$fit_besag2_g <- r1$summary.fitted.values[1:b+b,"mean"]/map_2$D.E

l <- leaflet(map_2) %>% addTiles()

pal <- colorNumeric(palette = "YlOrRd", 
                    domain = seq(min(map_2$fit_besag2_m)-0.5,max(map_2$fit_besag2_m)+0.5 , by = 0.1))

l %>% addPolygons(color = "grey", weight = 1, 
                  fillColor = ~pal(map_2$fit_besag2_m), 
                  fillOpacity = 0.5) %>%
  addLegend(pal = pal, values= ~seq(min(map_2$fit_besag2_m)-0.5,max(map_2$fit_besag2_m)+0.5 , by = 0.1), 
            opacity = 0.5, title = "RR from besag2 model for malaria",
            position = "topright")

pal <- colorNumeric(palette = "YlOrRd", 
                    domain = seq(min(map_2$fit_besag2_g)-0.5,max(map_2$fit_besag2_g)+0.5 , by = 0.1))


l %>% addPolygons(color = "grey", weight = 1, 
                  fillColor = ~pal(map_2$fit_besag2_g), 
                  fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~seq(min(map_2$fit_besag2_g)-0.5,max(map_2$fit_besag2_g)+0.5 , by = 0.1), 
            opacity = 0.5, title = "RR from besag2 model for G6PD",
            position = "topright")


#Common spatial field
map_2$besag_field <- r1$summary.random$ID$mean[1:b]*r1$summary.hyperpar["Scale paramter a for ID", "mean"]

pal <- colorNumeric(palette = "YlOrRd", 
                    domain = seq(min(map_2$besag_field)-0.5,max(map_2$besag_field)+0.5 , by = 0.1))

l %>% addPolygons(color = "grey", weight = 1, 
                  fillColor = ~pal(map_2$besag_field), 
                  fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~seq(min(map_2$besag_field)-0.5,max(map_2$besag_field)+0.5 , by = 0.1), 
            opacity = 0.5, title = "Common spatial field",
            position = "topright")


#Disease-specific BYM effects
map_2$bym2_m <- r1$summary.random$b1$mean[1:b] 
map_2$bym2_g <- r1$summary.random$b2$mean[1:b] 


pal <- colorNumeric(palette = "YlOrRd", 
                    domain = seq(min(map_2$bym2_m)-0.5,max(map_2$bym2_m)+0.5 , by = 0.1))

l %>% addPolygons(color = "grey", weight = 1, 
                  fillColor = ~pal(map_2$bym2_m), 
                  fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~seq(min(map_2$bym2_m)-0.5,max(map_2$bym2_m)+0.5 , by = 0.1), 
            opacity = 0.5, title = "BYM values for malaria",
            position = "topright")

pal <- colorNumeric(palette = "YlOrRd", 
                    domain = seq(min(map_2$bym2_g)-0.5,max(map_2$bym2_g)+0.5 , by = 0.1))


l %>% addPolygons(color = "grey", weight = 1, 
                  fillColor = ~pal(map_2$bym2_g), 
                  fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~seq(min(map_2$bym2_g)-0.5,max(map_2$bym2_g)+0.5 , by = 0.1), 
            opacity = 0.5, title = "BYM values for G6PD",
            position = "topright")


```

```{r, eval = F}

#Disease-specific Besag effects
map_2$bym2_m <- r1$summary.random$b1$mean[b+1:b] 
map_2$bym2_g <- r1$summary.random$b2$mean[b+1:b] 

pal <- colorNumeric(palette = "YlOrRd", 
                    domain = seq(min(map_2$bym2_m)-0.5,max(map_2$bym2_m)+0.5 , by = 0.1))

l %>% addPolygons(color = "grey", weight = 1, 
                  fillColor = ~pal(map_2$bym2_m), 
                  fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~seq(min(map_2$bym2_m)-0.5,max(map_2$bym2_m)+0.5 , by = 0.1), 
            opacity = 0.5, title = "Besag values for malaria",
            position = "topright")

pal <- colorNumeric(palette = "YlOrRd", 
                    domain = seq(min(map_2$bym2_g)-0.5,max(map_2$bym2_g)+0.5 , by = 0.1))


l %>% addPolygons(color = "grey", weight = 1, 
                  fillColor = ~pal(map_2$bym2_g), 
                  fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~seq(min(map_2$bym2_g)-0.5,max(map_2$bym2_g)+0.5 , by = 0.1), 
            opacity = 0.5, title = "Besag values for G6PD",
            position = "topright")

# the iid for malaria
map_2$vm = (r1$summary.random$b1$mean[1:21] * sqrt(r1$summary.hyperpar$mode[3]) -  
             r1$summary.random$b1$mean[22:42] * sqrt(r1$summary.hyperpar$mode[4])) / 
  sqrt(1 - r1$summary.hyperpar$mode[4])

range(map_2$vm)

pal <- colorNumeric(palette = "YlOrRd", domain = seq(min(map_2$vm)-0.2, max(map_2$vm) + 0.5 , by = 0.1))

l %>% addPolygons(color = "grey", weight = 1, 
                  fillColor = ~pal(map_2$vm), 
                  fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~seq(min(map_2$vm)-0.2, max(map_2$vm) + 0.5 , by = 0.1), 
            opacity = 0.5, title = "IID of Malaria",
            position = "topright")


# iid effect for g6pd
map_2$v = (r1$summary.random$b2$mean[1:21] * sqrt(r1$summary.hyperpar$mode[5]) -  
              r1$summary.random$b2$mean[22:42] * sqrt(r1$summary.hyperpar$mode[6])) / 
              sqrt(1 - r1$summary.hyperpar$mode[6])

pal <- colorNumeric(palette = "YlOrRd", domain = seq(min(map_2$v)-0.2, max(map_2$v) + 0.5 , by = 0.1))


l %>% addPolygons(color = "grey", weight = 1, 
                  fillColor = ~pal(map_2$v), 
                  fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~seq(min(map_2$v)-0.2, max(map_2$v) + 0.5 , by = 0.1), 
            opacity = 0.5, title = "IID for G6PD",
            position = "topright")



```
## Example 4 - Quantile models for disease mapping
Now we can also do quantile modeling, instead of the mean risk we can model the high risk areas only and see if there are different covariate impacts for the higher levels of risk. 

\begin{eqnarray}
y_{i 1}|\lambda_{i1} &\sim& \text { Poisson }\left(E_{i1}\lambda_{i 1}\right) \notag\\
y_{i 2}|\lambda_{i2} &\sim& \text { Poisson }\left(E_{i2}\lambda_{i 2}\right) \notag\\
\log (q_{i 1, \alpha_{1}}) &=& \eta_{i 1,\alpha_{1}} =  m_{1} + \sum_{f = 1}^{F_1} \beta_f X_{i f} + \sum_{r=1}^{R_1} \rho^r(u_{i r}) + b_{i 1} + S_{i} \notag \\
\log (q_{i 2,\alpha_{2}}) &=& \eta_{i 2,\alpha_{2}} = m_{2} + \sum_{f = 1}^{F_2} \gamma_f Z_{i f} + \sum_{r=1}^{R_2} \xi^r(v_{i r})+ b_{i 2}  + c \, S_{i},\label{qa2}
\end{eqnarray}. 
In INLA we only need to add the special link function as follows:  
control.family = list(control.link = list(model = "quantile", quantile = alpha)). 
                                                      
First we model the malaria incidence quantile and thereafter the G6PD incidence, while ignoring a shared effect.


```{r pre, eval = T}
alpha0 = 0.8
r.m <- inla(formula = map_2$M.cases ~ 1 + offset(log(map_2$M.E)) +
              f(b1, model = "bym2", scale.model = T,
                graph = g)   ,
            family = "poisson",
            data =  merg.data,
            control.predictor = list(compute = T),
            control.family = list(control.link = list(model = "quantile",
                                                      quantile = alpha0)),
            verbose = F)

round(r.m$summary.hyperpar[,c(1,3,5,6)],3)


```


```{r}
# the quantile for G6PD only
r.d <- inla(formula = map_2$D.cases ~ 1 + offset(log(map_2$D.E))+
              f(b2,model = "bym2", scale.model = T,
                graph = g)  ,
            family = "poisson",
            data =  merg.data,
            control.predictor = list(compute = T),
            control.family = list(control.link = list(model = "quantile",
                                                      quantile = alpha0)),
            verbose = F)

round(r.d$summary.hyperpar[,c(1,3,5,6)],3)
```
Visualize the effects.
```{r}
# the iid EFFECT malaria
map_2$vm2 = (r.m$summary.random$b1$mean[1:21] * sqrt(r.m$summary.hyperpar$mode[1]) -  
              r.m$summary.random$b1$mean[22:42] * sqrt(r.m$summary.hyperpar$mode[2])) / 
  sqrt(1 - r.m$summary.hyperpar$mode[2])

pal <- colorNumeric(palette = "YlOrRd", domain = seq(-2.6,1.6 , by = 0.1))


l %>% addPolygons(color = "grey", weight = 1, 
                  fillColor = ~pal(map_2$vm2), 
                  fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~seq(-2.6,1.6 , by = 0.1), 
            opacity = 0.5, title = "IID effect for G6PD",
            position = "topright")

# the iid EFFECT g6

map_2$vd2 = (r.d$summary.random$b2$mean[1:21] * sqrt(r.d$summary.hyperpar$mode[1]) -  
              r.d$summary.random$b2$mean[22:42] * sqrt(r.d$summary.hyperpar$mode[2])) / 
  sqrt(1 - r.d$summary.hyperpar$mode[2])

pal <- colorNumeric(palette = "YlOrRd", domain = seq(-5.6,2.6 , by = 0.1))


l %>% addPolygons(color = "grey", weight = 1, 
                  fillColor = ~pal(map_2$vd2), 
                  fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~seq(-5.6,2.6 , by = 0.1), 
            opacity = 0.5, title = "IID effect for G6PD",
            position = "topright")



# the spatial effect for malaria

map_2$u.m2 = r.m$summary.random$b1$mean[22:42]

pal <- colorNumeric(palette = "YlOrRd", domain = seq(-5.6,2.6 , by = 0.1))


l %>% addPolygons(color = "grey", weight = 1, 
                  fillColor = ~pal(map_2$u.m2), 
                  fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~seq(-5.6,2.6 , by = 0.1), 
            opacity = 0.5, title = "Besag effect for Malaria",
            position = "topright")

# the spatial effect for G6PD

map_2$u.d2 = r.d$summary.random$b2$mean[22:42]


pal <- colorNumeric(palette = "YlOrRd", domain =seq(-5.6,2.6 , by = 0.1))


l %>% addPolygons(color = "grey", weight = 1, 
                  fillColor = ~pal(map_2$u.d2), 
                  fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~ seq(-5.6,2.6 , by = 0.1), 
            opacity = 0.5, title = "Besag effect for G6PD",
            position = "topright")


# The relative risk for malaria 

map_2$rrm = r.m$summary.fitted.values[,"mean"]/map_2$M.E
range(map_2$rrm)

pal <- colorNumeric(palette = "YlOrRd", 
                    domain = seq(0,6.3 , by = 0.1))
l %>% addPolygons(color = "grey", weight = 1, 
                  fillColor = ~pal(map_2$rrm), 
                  fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~seq(0,6.3 , by = 0.1), 
            opacity = 0.5, title = "RR",
            position = "topright")


# The relative risk for G6PD

map_2$rrd = r.d$summary.fitted.values[,"mean"]/map_2$D.E
range(map_2$rrd)


pal <- colorNumeric(palette = "YlOrRd", domain =seq(0, 6.3, length = 9))


l %>% addPolygons(color = "grey", weight = 1, 
                  fillColor = ~pal(map_2$rrd), 
                  fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~ seq(0, 6.3, length = 9), 
            opacity = 0.5, title = "RR",
            position = "topright")

# The predicted cases for Malaria 

map_2$Predicted.MM = r.m$summary.fitted.values$mean

pal <- colorNumeric(palette = "YlOrRd", domain =seq(0,40000, length =9))


l %>% addPolygons(color = "grey", weight = 1, 
                  fillColor = ~pal(map_2$Predicted.MM), 
                  fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~ seq(0,40000, length =9), 
            opacity = 0.5, title = "Predicted",
            position = "topright")


#Observed

l %>% addPolygons(color = "grey", weight = 1, 
                  fillColor = ~pal(map_2$M.cases), 
                  fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~ seq(0,40000, length =9), 
            opacity = 0.5, title = "Observed",
            position = "topright")




# The predicted cases for G6PD


map_2$Predicted.DD = r.d$summary.fitted.values$mean
range(map_2$Predicted.DD)


pal <- colorNumeric(palette = "YlOrRd", domain =seq(0,500 , by = 10))


l %>% addPolygons(color = "grey", weight = 1, 
                  fillColor = ~pal(map_2$Predicted.DD), 
                  fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~ seq(0,500 , by = 10), 
            opacity = 0.5, title = "Predicted",
            position = "topright")


l %>% addPolygons(color = "grey", weight = 1, 
                  fillColor = ~pal(map_2$D.cases), 
                  fillOpacity = 0.5) %>%
  addLegend(pal = pal, values = ~ seq(0,500 , by = 10), 
            opacity = 0.5, title = "Observed",
            position = "topright")


```
Now we can do a multivariate quantile disease mapping of these two diseases.
```{r, eval = T}
#Use the dataset and formula we already created for the joint mean modeling

formula = yy ~ -1 + m + offset(log(E1))+ offset(log(E2))+
  f(ID, model = "besag2", graph=g, scale.model = T)+
  f(ID_copy, copy="ID", hyper = list(beta = list(fixed = TRUE)))+
  f(b1 , model = "bym2", graph=g, scale.model=TRUE)+
  f(b2, model = "bym2", graph=g, scale.model=TRUE)

alpha0 = 0.7
r1 <- inla(formula,
           family = c("poisson","poisson"),
           
           control.family = list(list(control.link = list(model = "quantile",quantile = alpha0)),
                                 
                                 list(control.link = list(model = "quantile",
                                                          quantile = alpha0))), 
           data = d,
           verbose = F,
           control.compute = list(dic = T, waic = T, cpo = T),
           control.predictor = list(compute = T))


round(r1$summary.fixed[,c(1:3,5)],3)

round(r1$summary.hyperpar[,c(1,3,5,6)],3)

#Opposite quantiles
r1a <- inla(formula,
           family = c("poisson","poisson"),
           control.family = list(list(control.link = list(model = "quantile",quantile = alpha0)),
                                 list(control.link = list(model = "quantile",
                                                          quantile = 1-alpha0))),
           data = d,
           verbose = F,
           control.compute = list(dic = T, waic = T, cpo = T),
           control.predictor = list(compute = T))


round(r1a$summary.fixed[,c(1:3,5)],3)

round(r1a$summary.hyperpar[,c(1,3,5,6)],3)

```
Maybe the *besag2* is too strict because of the restriction a>0. We can add more flexibility with copying a besagproper model component as follows:
```{r}
#High quantile of malaria and low quantile of G6PD
alpha0 = 0.7
d1 <- d
d1$ID_copy = c(rep(NA, b),1:b)

formula1 = yy ~ -1 + m + offset(log(E1))+ offset(log(E2))+
  f(ID, model = "besagproper", graph=g) +
  f(ID_copy, copy="ID", hyper = list(beta = list(fixed = FALSE))) +
  f(b1 , model = "bym2", graph=g, scale.model=TRUE)+
  f(b2, model = "bym2", graph=g, scale.model=TRUE)

r2a <- inla(formula1,
           family = c("poisson","poisson"),
           control.family = list(list(control.link = list(model = "quantile",quantile = alpha0)),
                                 list(control.link = list(model = "quantile",
                                                          quantile = 1-alpha0))),
           data = d1,
           verbose = F,
           control.compute = list(dic = T, waic = T, cpo = T),
           control.predictor = list(compute = T))


round(r2a$summary.fixed[,c(1:3,5)],3)

round(r2a$summary.hyperpar[,c(1,3,5,6)],3)

#Median joint modeling

alpha0 = 0.5

formula1 = yy ~ -1 + m + offset(log(E1))+ offset(log(E2))+
  f(ID, model = "besagproper", graph=g) +
  f(ID_copy, copy="ID", hyper = list(beta = list(fixed = FALSE))) +
  f(b1 , model = "bym2", graph=g, scale.model=TRUE)+
  f(b2, model = "bym2", graph=g, scale.model=TRUE)

r3a <- inla(formula1,
           family = c("poisson","poisson"),
           control.family = list(list(control.link = list(model = "quantile",quantile = alpha0)),
                                 list(control.link = list(model = "quantile",
                                                          quantile = 1-alpha0))),
           data = d1,
           verbose = F,
           control.compute = list(dic = T, waic = T, cpo = T),
           control.predictor = list(compute = T))


round(r3a$summary.fixed[,c(1:3,5)],3)

round(r3a$summary.hyperpar[,c(1,3,5,6)],3)

```



